Intrinsic Variables and State Reconstruction in Multiscale Simulations

### 2013-12-27 paper/draft3.tex
Added files related to paper

\documentclass[aip,jcp,reprint,twocolumn]{revtex4-1}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{mhchem}

@@

%\author{}
%\email{}
%\affiliation{}

@@

\begin{abstract}
Finding informative low-dimensional descriptions of high dimensional simulation data
(like the ones arising in molecular dynamics or kinetic Monte Carlo simulations of
physical/chemical processes) is crucial in understanding physical phenomena, and can
also dramatically assist in accelerating the simulations themselves.
%
In this paper, we discuss and illustrate the use of nonlinear intrinsic variables (NIVs)
in the mining of high-dimensional multiscale simulation data.
%
In particular, we focus on the way that NIVs allow us to functionally merge different
simulation ensembles, and {\em different partial observations of these ensembles}, as well
as to infer variables not explicitly measured.
%
The approach relies on certain simple features of the underlying process variability to
filter out measurement noise and systematically recover a unique reference coordinate frame.
%
We illustrate the approach through two distinct sets of multiscale atomistic simulations: a fast/slow
time scale simulation of an enzyme reaction network, and a molecular
dynamics simulation of alanine dipeptide in explicit water.

@@

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data.
%
These developments have the potential to significantly enable the computational exploration
of phyiscochemical problems.
%
If the (high-dimensional) data $Y(t)$ arise from, say, a
molecular dynamics simulation of a macromolecule in solution, or from the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $x(t)$ is invaluable in understanding and predicting system behavior.

%
While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
such data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis,
different data sets on the same low-dimensional hyperplane in the ambient space $Y$
will lead to different basis vectors - in effect, to different $x$.
%
While this can be easily fixed by a linear transformation, the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data ensembles are obtained using different instrumental modalities
(say, when one wants to merge molecular dynamics data with spectral information).
%
Clearly, the ability to systematically construct a {\em unique, consistent} reduction
coordinate set, shared by all measurement ensembles and observation modalities -
what we call a (Nonlinear) Intrinsic Variable Set - is invaluable.
%
Embedding data in such a coordinate system allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements through the construction
of nonlinear observers of unmeasured variables.
%
This latter empirical mapping requires good interpolation tools in embedding space; to this
end we will demonstrate here the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
%
We chose two distinct multiscale illustrative examples. The first is a simulation
of two Goldbeter-Koshland modules in an enzyme kinetics model \cite{zagaris2012stability} using Gillespie's Stochastic Simulation
Algorithm (SSA); in certain parameter regimes, separation of time scales is known
to reduce this model to an effective two-dimensional description.
%
Although this example is rather simple, it will serve as a nice introduction to our techniques and highlight the main features of the algorithms.
%
The second example is a molecular dynamics
simulation (in explicit water) of a simple peptide fragment (alanine dipeptide) whose folding
dynamics are known to be described through a small set of physical observables.
%
This example will demonstrate how our techniques outperform more common techniques, such as diffusion maps for dimensionality reduction
and nearest neighbor interpolation for reconstruction.
%
The paper is structured as follows: in Section \ref{sec:NIV} we present the Nonlinear Intrinsic Variable formulation and
the associated inference method. 
%
Section \ref{sec:LapPyr} contains our discussion of Laplacian Pyramids that
will be used for the completion of partial observations. 
%
Section \ref{sec:examples} contain the results
of applying the approach to simulation data from our two illustrative examples. 
%
We conclude with
a summary and discussion of open issues in Section \ref{sec:conclusions}.

@@

\subsection{Overview}
Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
The measured process is assumed to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
\begin{equation}
	\mathbf{Y}(t) = f(\mathbf{x}(t))
\end{equation}
where $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
%
The dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
\begin{equation}
	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
\end{equation}
where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises TODO: should there be a dot here?.

Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$. 
%
By relying on the fact that the empirical method we will describe is independent of the observation function $f$, we present the notion of intrinsic modeling and refer to the coordinates of $\mathbf{x}(t)$ as a Nonlinear Intrinsic Variable (NIV) set.
%
The available samples $\mathbf{Y}(t)$ may be the product of different measurement functions $f$ in various observable domains, or they maybe partial measurements consisting of merely a subset of the coordinates of the observable domains.
%
The idea would be to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation modality (see Figure \ref{fig:IntrinsicIllustration} for illustration).

The presented method consists of the following main principles. 
%
(1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise, and hence, creates a ``sphere" of samples in the underlying domain. 
%
This sphere is mapped to an ellipse in the observable domain by the measurement function $f$. 
%
In this work, the identification of the associated ellipse of samples according to the time labels enables to explore and learn the tangent planes of the observable manifolds via the principal components of the covariance matrices of the samples in these ellipses. 
%
(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be locally invariant to the measurement function $f$. 
%
(3) The NIV are constructed through the eigenvalue decomposition of a Laplace operator that is built upon a pairwise affinity between the samples, defined using the Riemannian metric.

\begin{figure*}[ht]
\includegraphics[width=0.7\textwidth]{IntrinsicEmbeddingIllustration2}
\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement modality. (Bottom) The underlying variables in which the diffusion parameters are independent with unit variance. The circle illustrates successive sample in a short burst/trajectory in time that create a sphere on the parametric manifold. (Top Left) The first set of observed variables. The ellipse illustrates the mapping of the sphere of the underlying samples into the observable domain via the first observation modality. In this domain, we illustrate that the observations might be partial, i.e., consist of merely a subset of the observed domain variables. (Top Right) The second set of observable variables. The ellipse illustrates the mapping of the sphere of the underlying samples into a different observable domain via a second observation modality.}
\label{fig:IntrinsicIllustration}
\end{figure*}

@@

In particular, we define a Riemannian metric between a pair of samples using the associated covariance matrices as
\begin{align}
	d(\mathbf{Y}(t), \mathbf{Y}(\tau)) &= 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1} \nonumber \\
	&\times (\mathbf{Y}(t) - \mathbf{Y}(\tau)).
	\label{eq:mahalanobis}
\end{align}
This metric \eqref{eq:mahalanobis} is called the Mahalanobis distance.
%
As previously described, the covariance matrices convey the local variability of the measurements and are utilized to explore and learn the tangent planes of the observable manifold.
%
This information is then utilized in \eqref{eq:mahalanobis} to compare a pair of points according to the directions of their respected tangent planes.

The Mahalanobis distance is invariant under linear transformations.
%
Thus, by assuming that the observation function $f$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $f(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to a second order, i.e.,
\begin{equation}
	\| \mathbf{x}(t) - \mathbf{x}(\tau) \|^2 = d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) + \mathcal{O}(\| \mathbf{Y}(t) - \mathbf{Y}(\tau)\|^4).
\end{equation}
%
This result implies that the Mahalanobis distance is invariant to the measurement function $f$, and hence, 
it yields the same distances between samples obtained under different observation modalities or even partial observations.

\subsection{Laplace Operator}
The Mahalanobis distance described in Section \ref{subsec:mahalanobis} enables to compare the observations in terms of the intrinsic variables of the associated underlying diffusion process.
%
In this section, we recover the underlying process itself from the pairwise Euclidean distances through the eigenvectors of a Laplace operator.

Let $\mathbf{W}$ be a pairwise affinity matrix (kernel) based on a Gaussian, whose $(t,\tau)$-th element is given by
\begin{equation}
	W_{t,\tau} = \exp \left\{ - \frac{ d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) )} {\varepsilon}\right\}
	\label{eq:kernel}
\end{equation}
where $\varepsilon$ is the kernel scale, which can be set according to \cite{coifman2008graph}.
%
Let $\mathbf{D}$ be a diagonal matrix whose diagonal elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$ be a normalized kernel that shares its eigenvectors with the normalize graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}}$ \cite{chung1997spectral}. 
%
These eigenvectors $\psi_j$ reveal the underlying structure of the data TODO:cite. 
%
Specifically, the $i$-th coordinate of the $j$-th eigenvector can be associated with an intrinsic coordinate $j$ of the sample $\mathbf{x}(i)$ of the underlying process. 
%
The above steps to construct the nonlinear intrinsic variables are summarized in Algorithm \ref{algo}.

%TODO: do we need to describe the extension procedure?! - the recovery of the intrinsic variables through the eigenvalue decomposition (EVD) of the Laplace operator is computationally heavy...\\
TODO: we can discuss the probabilistic interpretation and explicitly state that we set the kernel scale to the median of the kernel.\\

Each eigenvector then describes an intrinsic variable for the data set of interest.
%
Because eigenvectors can be determined up to a scaling factor, we must scale the eigenvectors from different data sets in a way so that the resulting embeddings are consistent.
%
We first scale the eigenvectors so that the norm of the each eigenvector is equal to the cardinality of the data set.
%
Intuitively, a data set that has twice as many samples as another data set, but the same underlying density and shape, should have twice the norm.
%
Still, the embedding eigenevctors for two identical data sets could differ by a sign.
%
Reconciling the signs for the embeddings of different data sets can be done in several ways and is somewhat problem-specific.
%
For example, if the mean of the embedding is sufficiently far from 0, the signs can be chosen such that the mean of each eigenvector is greater than 0;
alternatively, we can construct the data sets such that a few points are common between both data sets, and the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of these common points.

@@

Laplacian Pyramids (LP) is a multi-scale algorithm for approximating and extending an empirical function $f$ defined on a set of points.
%
The algorithm uses Laplacian kernels of decreasing widths to create multi-scale representations of $f$.
%
These representations can be easily extended to new data points.
%
This type of multi-scale representation was introduced by Burt and Adelson \cite{burt1983laplacian} for image coding,
and was later shown to be a tight frame by Do and Veterli \cite{do2003framing}.
%
Recently, LP was used to extend nonlinear embedding coordinates to new high-dimensional data points \cite{rabin2012heterogeneous}.
%
We will first review the LP algorithm for approximating and extending a one-dimensional function.
%
We will then describe the application of LP in the context of this work - extending a high-dimensional function that is known on a set of points which lie in a low-dimensional space.

Let $f: \Gamma \rightarrow \mathbb{R}^d$ be a function that is known on a subset of points $S \subset \Gamma$.
%
A coarse representation of $f$ is generated using a coarse smoothing operator $P_0$.
%
The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by $p_0(t,\tau)= s_0^{-1}(t)w_0(t,\tau)$, where $w_0(t,\tau)=e^{-d^2(t,\tau) / \sigma_0}$. 
%
The pairwise distance $d(t,\tau)$ is typically the Euclidian distance. 
%
The parameter $\sigma_0$ is set to be large compared to the values of $d(t,\tau)$  and $s_0^{-1}(t)=\sum_{\tau \in S}w_0(t,\tau)$ is the normalizing term. 
%
The application of $P_0$ to $f$ yields a coarse representation of the function, which we denote by $f_0=P_0(f)$.

The difference $\delta_1 = f-P_0(f)$ is input for the next iteration of the algorithm, 
which uses the smoothing operator $P_1$, $P_1 \propto e^{-d^2(t,\tau)^2 / 2^{-1} \sigma_0}$, to construct a coarse representation of $\delta_1$.
%
The obtained  representation of $\delta_1$, $P_1(\delta_1)$ together with the result of the previous iteration $f_0 = P_0(f)$ 
yields a new, finer representation of $f$, $f_1 = P_0(f) + P_1(\delta_1)$.

In an iterative manner, multi-scales representations of the function $f$, denoted by $f_l$ are constructed (see Eq. \eqref{eq:LP_multi_scale}). 
%
As $l$ grows, the approximation is of a finer resolution because $P_l \propto e^{-d^2(t,\tau) / 2^{-l} \sigma_0}$ uses a Laplacian kernel of a finer width. 
%
The iterations stop when the difference between $f$ and $f_l$ is smaller than a pre-defined error threshold.

\begin{equation} \label{eq:LP_multi_scale}
 \begin{array}{cl}
\mbox{scale 0:} & f_0 = P_0(f) \\
\mbox{scale 1:} & f_1 = P_0(f) + P_1(\delta_1) \\
: & : \\
\mbox{scale l:} & f_l = P_0(f) + \sum_{k=1}^{l}P_k(\delta_k)\\
\end{array}
\end{equation}

The representations $f_0$, $f_1$,..$f_l$ can be extended to a new point $\tilde{t} \in \Gamma \backslash S $ by extending the operators $P_0, P_1,\ldots,P_l$. 
%
For example, $f_0(\tilde{t}) = \sum_{t \in S} p_0(\tilde{t}, t)f(t)$ and
$f_1(\tilde{t}) = f_0(\tilde{t}) + \sum_{(t \in S)}p_1(\tilde{t}, t)\delta_1(t)$.

\begin{widetext}
Figure \ref{fig:LP_ex} displays an illustrative example of the algorithm, when applied to the function 
 \begin{equation} \label{eq:LP_example}
f(x) = \left\{
\begin{array}{l l}
-0.02(x-4\pi)^2 + \sin(x) &  0 \le x \le 4\pi \\
-0.02(x-4\pi)^2 + \sin(x) + 0.5 \sin(3x) &  4\pi < x \le 7.5 \pi \\
-0.02(x-4\pi)^2 + \sin(x) + 0.5 \sin(3x) +0.25 \sin(9x) &  7.5 \pi < x \le 10 \pi
\end{array}
\right.
\end{equation}
that contains several scales. 
%
The coarse regions of the function ( $0 \le x \le 4\pi$ ) are well approximated by a small number of scales. 
%
As the function becomes more oscillatory ( $\pi \le x \le 7.5\pi$ and $7.5\pi \le x \le 10\pi$), 
a finer representation, (meaning a larger number of scales  $l$ )  is required to capture its behavior.
\end{widetext}

>>>>

\documentclass[aip,jcp,preprint]{revtex4-1}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{natbib}
%\usepackage{mhchem}

@@

\author{Carmeline J. Dsilva}
\email{cdsilva@princeton.edu}
\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA}

\author{Ronen Talmon}
\email{ronen.talmon@yale.edu}
\affiliation{Department of Mathematics, Yale University, New Haven, Connecticut, USA}

\author{Neta Rabin}
\email{netar@afeka.ac.ilil}
\affiliation{Department of Exact Sciences, Afeka Tel-Aviv Academic College of Engineering, Tel-Aviv, Israel}

\author{Ronald R. Coifman}
\email{coifman@math.yale.edu}
\affiliation{Department of Mathematics, Yale University, New Haven, Connecticut, USA}

\author{Ioannis G. Kevrekidis}
\email{yannis@princeton.edu}
\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA}
\affiliation{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}

@@

\begin{abstract}
Finding informative low-dimensional descriptions of high-dimensional simulation data
(like the ones arising in molecular dynamics or kinetic Monte Carlo simulations of
physical and chemical processes) is crucial to understanding physical phenomena, and can
also dramatically assist in accelerating the simulations themselves.
%
In this paper, we discuss and illustrate the use of nonlinear intrinsic variables (NIV)
in the mining of high-dimensional multiscale simulation data.
%
In particular, we focus on the way NIV allow us to functionally merge different
simulation ensembles, and {\em different partial observations of these ensembles}, as well
as to infer variables not explicitly measured.
%
The approach relies on certain simple features of the underlying process variability to
filter out measurement noise and systematically recover a unique reference coordinate frame.
%
We illustrate the approach through two distinct sets of multiscale atomistic simulations:
a simulation of an enzyme reaction network exhibiting both fats and slow time scales,
 and a molecular dynamics simulation of alanine dipeptide in explicit water.

@@

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data.
%
These developments have the potential to significantly enable the computational exploration
of physicochemical problems.
%
If the (high-dimensional) data $\mathbf{Y}(t)$ arise from, for example, a
molecular dynamics simulation of a macromolecule in solution or the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $\mathbf{x}(t)$ is invaluable in understanding and predicting system behavior.

%
While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
such data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis,
different data sets on the same low-dimensional hyperplane in the ambient space
will lead to different basis vectors - in effect, to different reduction coordinates $\mathbf{x}$.
%
While this can be easily rectified by a linear transformation, the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data sets are obtained using different instrumental modalities
(such as when one wants to merge molecular dynamics data with spectral information).
%
Clearly, the ability to systematically construct a {\em unique, consistent} reduction
coordinate set, shared by all measurement ensembles and observation modalities -
what we call Nonlinear Intrinsic Variables - is invaluable.
%
Embedding data in such a coordinate system allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements through the construction
of nonlinear observers of unmeasured variables.
%
This empirical mapping requires good interpolation tools in embedding space; to this
end, here we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.

We will demonstrate our methodologies with two distinct illustrative examples.
%
The first is a simulation
of two Goldbeter-Koshland modules in an enzyme kinetics model \cite{zagaris2012stability} using Gillespie's Stochastic Simulation
Algorithm (SSA); in certain parameter regimes, separation of time scales is known
to reduce this model to an effective two-dimensional description.
%
Although this example is rather simple, it will serve as a nice introduction to our techniques and highlight the main features of the algorithms.
%
The second example is a molecular dynamics
simulation (in explicit water) of a simple peptide fragment (alanine dipeptide) whose folding
dynamics are known to be described through a small set of physical observables.
%
This example will demonstrate how our techniques outperform more common techniques,
such as diffusion maps for dimensionality reduction
and nearest neighbor interpolation for interpolation.
%
The paper is structured as follows: in Section \ref{sec:NIV} we present the Nonlinear Intrinsic Variable formulation and
the associated inference method.
%
Section \ref{sec:LapPyr} contains our discussion of Laplacian Pyramids that
will be used for the completion of partial observations.
%
Section \ref{sec:examples} contain the results
of applying the approach to simulation data from our two illustrative examples.
%
We conclude with
a summary and discussion of open issues in Section \ref{sec:conclusions}.

@@

\subsection{Overview}
Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
The measured process is assumed to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
\begin{equation}
	\mathbf{Y}(t) = f(\mathbf{x}(t))
\end{equation}
where $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
%
The dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
\begin{equation}
	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
\end{equation}
where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises.

Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$.
%
By relying on the fact that the empirical method we will describe is independent of the observation function $f$, we present the notion of intrinsic modeling and refer to the coordinates of $\mathbf{x}(t)$ as a Nonlinear Intrinsic Variable (NIV) set.
%
The available samples $\mathbf{Y}(t)$ may be the product of different measurement functions $f$ in various observable domains, or they maybe partial measurements consisting of merely a subset of the coordinates of the observable domains.
%
The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $f$ (see Figure \ref{fig:IntrinsicIllustration} for illustration).

The presented method consists of the following main principles.
%
(1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise, and hence, creates a ``sphere" of samples in the underlying domain.
%
This sphere is mapped to an ellipse in the observable domain by the measurement function $f$.
%
In this work, the identification of the associated ellipse of samples according to the time labels enables us
to estimate the tangent planes of the observable manifolds via the principal components of the covariance matrices of the samples in these ellipses.
%
(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be locally invariant to the measurement function $f$.
%
(3) The NIV are constructed through the eigenvalue decomposition of a Laplace operator that is built upon a pairwise affinity between the samples, defined using the Riemannian metric.

\begin{figure*}[ht]
\includegraphics[width=0.7\textwidth]{IntrinsicEmbeddingIllustration2}
\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement function $f$. (Bottom) The underlying variables in which the coordinates are independent with unit variance. The circle illustrates samples in a short trajectory in time that create a sphere on the manifold. (Top Left) The first set of observed variables. The ellipse illustrates the mapping of the sphere of the underlying samples into the observable domain via the first observation function. In this domain, we illustrate that the observations might be partial, i.e., consist of merely a subset of the observed domain variables. (Top Right) The second set of observable variables. The ellipse illustrates the mapping of the sphere of the underlying samples into a different observable domain via a second observation function.}
\label{fig:IntrinsicIllustration}
\end{figure*}

@@

In particular, we define a Riemannian metric between a pair of samples using the associated covariance matrices as
\begin{equation}
	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1}(\mathbf{Y}(t) - \mathbf{Y}(\tau)).
	\label{eq:mahalanobis}
\end{equation}
This metric \eqref{eq:mahalanobis} is called the Mahalanobis distance.
%
As previously described, the covariance matrices convey the local variability of the measurements and are utilized to explore and learn the tangent planes of the observable manifold.
%
This information is then utilized in \eqref{eq:mahalanobis} to compare a pair of points according to the directions of their respected tangent planes.

The Mahalanobis distance is invariant under linear transformations.
%
Thus, by assuming that the observation function $f$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $f(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to a second order, i.e.,
\begin{equation}
	\| \mathbf{x}(t) - \mathbf{x}(\tau) \|^2 = d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) + \mathcal{O}(\| \mathbf{Y}(t) - \mathbf{Y}(\tau)\|^4).
\end{equation}
%
This result implies that the Mahalanobis distance is invariant to the measurement function $f$, and hence,
it yields the same distances between samples obtained under different observation functions or even partial observations.

\subsection{Laplace Operator}
The Mahalanobis distance described in Section \ref{subsec:mahalanobis} enables us to compare observations in terms of the intrinsic variables of the associated underlying diffusion process.
%
In this section, we recover the underlying process itself from the pairwise Euclidean distances through the eigenvectors of a Laplace operator.

Let $\mathbf{W}$ be a pairwise affinity matrix (kernel) based on a Gaussian, whose $(t,\tau)$-th element is given by
\begin{equation}
	W_{t,\tau} = \exp \left\{ - \frac{ d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) )} {\varepsilon}\right\}
	\label{eq:kernel}
\end{equation}
where $\varepsilon$ is the kernel scale, which can be set according to \onlinecite{coifman2008graph, hein2005intrinsic}.
%
Based on the kernel, we form a weighted graph, where the measurements $\mathbf{Y}(t)$ are the graph nodes and the weight of the edge connecting node $\mathbf{Y}(t)$ to node $\mathbf{Y}(\tau)$ is $W_{t,\tau}$. 
%
In particular, a Gaussian kernel exhibits a notion of locality by defining a neighborhood around each measurement $\mathbf{Y}(t)$ of radius $\varepsilon$, 
i.e., measurements $\mathbf{Y}(\tau)$ s.t. $d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) ) > \varepsilon$ are weakly connected to $\mathbf{Y}(t)$. 
%
In practice, we set $\varepsilon$ to be the median of the elements of the kernel matrix. 
%
According to the graph interpretation, it implies a well connected graph because each measurement is effectively connected to half of the other measurements.

Let $\mathbf{D}$ be a diagonal matrix whose diagonal elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
be a normalized kernel that shares its eigenvectors with the normalize graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}}$ \cite{chung1997spectral}.
%
The eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, denoted $\psi_j$, reveal the underlying structure of the data \cite{coifman2005geometric}.
%
Specifically, the $i$-th coordinate of the $j$-th eigenvector can be associated with an intrinsic coordinate $j$ of the sample $\mathbf{x}(i)$ of the underlying process.
%
The eigenvectors are ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$, where $\lambda_j$ is the eigenvalue associated with eigenvector $\psi_j$. 
%
Because $\mathbf{W}^{\mathrm{norm}} \sim \mathbf{D}^{-1}\mathbf{W} $, and $\mathbf{D}^{-1}\mathbf{W}$ is row-stochastic, 
$\lambda_1 = 1$ and $\psi_1$ is the diagonal of $\mathbf{D}^{1/2}$. 
%
The next few eigenvectors describe the geometry of the underlying manifold.
%
The above steps to construct the nonlinear intrinsic variables are summarized in Algorithm \ref{algo}.

Each eigenvector then describes an intrinsic variable for the data set of interest.
%
Because eigenvectors can be determined up to a scaling factor, we must scale the eigenvectors from different data sets such that the resulting embeddings are consistent.
%
We first scale the eigenvectors so that the norm of the each eigenvector is equal to the cardinality of the data set.
%
Intuitively, a data set that has twice as many samples as another data set, but the same underlying geometry, should have twice the norm.
%
Still, the embedding eigenevctors for two identical data sets could differ by a sign.
%
Reconciling the signs for the embeddings of different data sets can be done in several ways and is somewhat problem-specific.
%
For example, if the mean of the embedding is sufficiently far from 0, the signs can be chosen such that the mean of each eigenvector is greater than 0;
alternatively, we can construct the data sets such that a few points are common between both data sets, and the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of these common points.

@@

Laplacian Pyramids (LP) is a multiscale algorithm for approximating and extending an empirical function $f$ defined on a set of points.
%
The algorithm uses Laplacian kernels of decreasing widths to create multiscale representations of $f$;
These representations can be easily extended to new data points.
%
This type of multiscale representation was introduced by Burt and Adelson \cite{burt1983laplacian} for image coding,
and was later shown to be a tight frame by Do and Veterli \cite{do2003framing}.
%
Recently, LP was used to extend nonlinear embedding coordinates to new high-dimensional data points \cite{rabin2012heterogeneous}.
%
We will first review the LP algorithm for approximating and extending a one-dimensional function,
and then describe the application of LP in the context of this work - extending a high-dimensional function that is known on a set of points which lie in a low-dimensional space.

Let $f: \Gamma \rightarrow \mathbb{R}^n$ be a function that is known on a subset of points $S \subset \Gamma$.
%
A coarse representation of $f$ is generated using a coarse smoothing operator $P_0$.
%
The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by $p_0(t,\tau)= s_0^{-1}(t)w_0(t,\tau)$,
where $w_0(t,\tau)=e^{-d^2(t,\tau) / \sigma_0}$ and $s_0(t)=\sum_{\tau \in S}w_0(t,\tau)$ is the normalizing term.
%
The pairwise distance $d(t,\tau)$ is typically the Euclidian distance, and the parameter $\sigma_0$ is set to be large compared to the values of $d^2(t,\tau)$.
%
The application of $P_0$ to $f$ yields a coarse representation of the function, which we denote by $f_0=P_0(f)$.

The difference $\delta_1 = f-P_0(f)$ is the input for the next iteration of the algorithm,
which uses the smoothing operator $P_1$, $P_1 \propto e^{-d^2(t,\tau)^2 / 2^{-1} \sigma_0}$, to construct a coarse representation of $\delta_1$.
%
The obtained representation of $\delta_1$, $P_1(\delta_1)$ together with the result of the previous iteration $f_0 = P_0(f)$
yields a new, finer representation of $f$, $f_1 = P_0(f) + P_1(\delta_1)$.
%
In an iterative manner, multiscale representations of the function $f$, denoted $f_l$, are constructed.
\begin{equation} \label{eq:LP_multi_scale}
 \begin{array}{cl}
\mbox{scale 0:} & f_0 = P_0(f) \\
\mbox{scale 1:} & f_1 = P_0(f) + P_1(\delta_1) \\
: & : \\
\mbox{scale l:} & f_l = P_0(f) + \sum_{k=1}^{l}P_k(\delta_k)\\
\end{array}
\end{equation}
As $l$ increases, the approximation becomes more refined because $P_l \propto e^{-d^2(t,\tau) / 2^{-l} \sigma_0}$ uses a Laplacian kernel of a finer width.
%
The iterations stop when the difference between $f$ and $f_l$ is smaller than a pre-defined error threshold.

The representations $f_0, f_1, \dots, f_l$ can be extended to a new point $\tilde{t} \in \Gamma \backslash S $ by extending the operators $P_0, P_1,\ldots,P_l$.
%
For example, $f_0(\tilde{t}) = \sum_{t \in S} p_0(\tilde{t}, t)f(t)$ and
$f_1(\tilde{t}) = f_0(\tilde{t}) + \sum_{(t \in S)}p_1(\tilde{t}, t)\delta_1(t)$.

\begin{widetext}
Figure \ref{fig:LP_ex} displays an illustrative example of the algorithm, when applied to the function
 \begin{equation} \label{eq:LP_example}
f(x) = \left\{
\begin{array}{l l}
-0.02(x-4\pi)^2 + \sin(x) &  0 \le x \le 4\pi \\
-0.02(x-4\pi)^2 + \sin(x) + 0.5 \sin(3x) &  4\pi < x \le 7.5 \pi \\
-0.02(x-4\pi)^2 + \sin(x) + 0.5 \sin(3x) +0.25 \sin(9x) &  7.5 \pi < x \le 10 \pi
\end{array}
\right.
\end{equation}
that contains several scales.
%
The coarse regions of the function ($0 \le x \le 4\pi$) are well approximated by a small number of scales.
%
As the function becomes more oscillatory ($\pi \le x \le 7.5\pi$ and $7.5\pi \le x \le 10\pi$),
a finer representation (meaning a larger number of scales  $l$) is required to capture its behavior.
\end{widetext}

In this work, LP is applied to extend a high-dimensional function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$, which maps a set of points in the NIV space to their values $\mathbf{Y}(t)$ in the observable space.
%
Let $\Psi(t) = \left(\psi_1(t),\psi_2(t),\ldots,\psi_d(t)\right)$ be the set of NIV that were constructed from the data samples $\mathbf{Y}(t)$, as described in Algorithm \ref{algo}.
%
The values of function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ are known on the subset $S = \{\Psi(t)\}$, with $f(\Psi(t)) = Y(t)$.
